Combining information from photometric and spectroscopic data:\\ Probabilistic spectrophotometric inference

### 2015-07-08 tex/ms.tex
more figure code

THIS SECTION NEEDS WORK; THIS IS JUST AN EQUATION DUMP.

HOGG: A model is a parameterized likelihood function, with prior PDFs
on the nuisance parameters (at least)...  There is a likelihood for
the photometry and for the spectroscopy...  Each of these penalizes
the difference between a column vector of data and a column vector of
model expectation value...

@@

{\color{blue} BDJ This paragraph should explain why we are using the
GP, not just how it works.}
For the spectroscopic measurements we are going to account for the
spectrophotometric calibration in our likelihood function.  We take
this calibration-vector that modifies the mean spectroscopic model to
produce the observed spectrum to be multiplicative with respect to the
mean model.  The multiplicative calibration vector can be explicitly
modeled, for example using polynomials.  However, for complex
calibration vectors over large wavelength ranges, the number of
parameters required to model the calibration can become quite large
(e.g. very high order polynomials) and the correct parameterization
may be unknown.  It is therefore desirable to introduce a
non-parameteric model of the calibration.  We do this using a Gaussian
Process: BDJ: prior over functions, analytically marginalizable,
covariant fractional uncertainties the same as multiplying by a
function (covariant uncertainties the same as adding a function.)
Using a polynomial can result in degeneracies between model and
polynomial parameters, whereas GP is only for things that {\it can't}
be fit by model.

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF_v2}
\ln p(d\given\theta,\alpha) &=&
                                -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta
                                + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - \mu(\theta) \ast f(\alpha) 
\\
C_{nn'} &\equiv& [\sigma_n^2 + (\hat{s} \, \mu)^2]\,\delta_{nn'} +
 K_\alpha(\lambda_n - \lambda_{n'}) \ast \mu \, \transpose{\mu}
\quad ,
\end{eqnarray}
where $\ast$ denotes element-wise multiplication,
$\theta$ are the physical parameters,
$\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelength scale and
amplitude of the residual calibration issues,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The multiplication of $\hat{s}$ and $K_\alpha$ by $\mu$ and
$\mu\,\transpose{\mu}$ respectively in the construction of $C_{nn'}$
signifies that they are defined in \emph{fractional} terms.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian Process is going to
pick up smaller-scale residuals away from \emph{that}.
The flexibility of the Gaussian Process specified by $C$ will obviate
the need for numerous high order polynomial terms in $f(\alpha)$ and
thus reduce the number of free parameters required in the model.  The
prior over functions implied by $C$ is analytically conditioned on the
residuals $\Delta$.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In detail, we are going to choose the exponential of a $M$th-order
polynomial form for the calibration vector $f(\alpha)$ and a bell
shape (Gaussian) for the kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& \exp\left(\sum_{m=o}^M c_m\,x^m\right)
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& a^2\,\exp\left(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2}\right)
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths
$\lambda_m$ and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.  The exponentiation of
the polynomial in ensures that $f(\alpha)$ is everywhere positive.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise.

\section{Data and Experiments}

We are now left to demonstrate that, by including photometry as well
as spectroscopy, it is possible to infer the parameters of interest
$\theta$ while marginalizing over the calibration parameters $\alpha$.
We will apply the methodology for spectrophotometric calibration
described above to both \emph{mock} and \emph{real} optical spectra of
Galactic globular clusters.  Spectra of globular clusters provide a
good test case, as they are less complex than galaxy spectra (for
which it is necessary to model the star formation history) but are by
no means trivial. In this way the key features of the calibration
model and methodology are not obscured.  Additionally, information
about the physical parameters of the cluster is available from
resolved stellar CMDs, providing an important check on our inference
methodology and especially our models for star cluster spectra.  However,
our methodology is applicable any spectral model: stars, galaxies or
ISM emission.

@@

Optical photometry is available from \citep{vanderbeke14} based on
SDSS for objects in the northern hemisphere and on data from CTIO for
southern hemisphere objects.  Briefly, total fluxes within the half-light radii
of the clusters \citet{harris98} were calculated, with small
corrections for stars unlikely to be cluster members based on their
positions in the CMD. For this data the formal Poisson uncertainties
are typically exceedingly small.  The dominant source of uncertainty
comes from the photometric calibration and from uncertainties in
cluster membership.

Near-Infrared photometry is available from \citet{cohen07}, who fit
King profiles \citet{king65} to 2MASS data in the $J$, $H$, and $K_s$
bands.  To produce NIR photometry consistent with the optical we
integrated the King profiles to the half-light radii used by
\citet{vanderbeke14}.

@@

The spectrum of one solar mass of stars that formed at lookback time
$t$ with metallicity $Z$ is given by $s(\lambda, t, Z [, \beta])$.
Here the additional parameters $\beta$ may include details of the
population synthesis model, such as the IMF slope and adjustments to
the stellar spectra or stellar evolutionary tracks. We will assume
$\beta$ to be fixed for the purposes of this exercise, with no loss of
generality but significant savings in computational time.  We will
also fix the shape of the attenuation curve, $k_\lambda $, to the
Milky way average of \citet{CCM89} with $R_V=3.1$ and assume that
$d_L$ is known to high precision from independent methods and can
therefore be treated as fixed.  In practice, $d_L$ is perfectly
degenerate with $m_*$. We calculate $s(\lambda, t, Z)$ using the
Flexible Stellar Population Synthesis code \citep[FSPS][]{fsps}.  We
use the Padova 2007 isochrones \citep{girardi00, bertilli94, marigo07}
and the MILES stellar spectral library \citep{miles_I, miles_II,
miles_III}.

@@

@@

\bibitem[Lee et al.(2011)]{xray_calunc} Lee, H., Kashyap, V.~L., 
van Dyk, D.~A., et al.\ 2011, \apj, 731, 126

@@

>>>>

A model is a parameterized likelihood function, with prior PDFs
on the nuisance parameters (at least).  There is a likelihood for
the photometry and for the spectroscopy...  Each of these penalizes
the difference between a column vector of data and a column vector of
model expectation value.

@@

{\color{blue} BDJ This paragraph should explain why we are using the
GP, not just how it works.}
For the spectroscopic measurements we are going to account for the
spectrophotometric calibration in our likelihood function.  We take
this calibration-vector that modifies the mean spectroscopic model to
produce the observed spectrum to be multiplicative with respect to the
mean model.  The multiplicative calibration vector can be explicitly
modeled, for example using polynomials.  However, for complex
calibration vectors over large wavelength ranges, the number of
parameters required to model the calibration can become quite large
(e.g. very high order polynomials) and the correct parameterization
may be unknown.  It is therefore desirable to introduce a
non-parameteric model of the calibration.  We do this using a Gaussian
Process: 
BDJ: prior over functions, analytically marginalizable, covariant
fractional uncertainties the same as multiplying by a function
(covariant absolute uncertainties the same as adding a function.)
Using a polynomial can result in degeneracies between model and
polynomial parameters, whereas GP only reduces the penalty for
residuals that {\it can't} be fit by the mean model.

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF_v2}
\ln p(d\given\theta,\alpha) &=&
                                -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta
                                + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - \mu(\theta) \ast f(\alpha) 
\\
C_{nn'} &\equiv& [\sigma_n^2 + (\hat{s} \, \mu_n)^2]\,\delta_{nn'} +
 K_\alpha(\lambda_n - \lambda_{n'}) 
\quad ,
\end{eqnarray}
where $\ast$ denotes element-wise multiplication,
$\theta$ are the physical parameters,
$\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelength scale and
amplitude of the residual calibration issues,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian Process is going to
pick up smaller-scale residuals away from \emph{that}.
The flexibility of the Gaussian Process specified by $C$ will obviate
the need for numerous high order polynomial terms in $f(\alpha)$ and
thus reduce the number of free parameters required in the model.  The
prior over functions implied by $C$ is analytically conditioned on the
residuals $\Delta$.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In detail, we are going to choose the exponential of a $M$th-order
polynomial form for the calibration vector $f(\alpha)$ and a bell
shape (Gaussian) for the kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& \exp\left(\sum_{m=o}^M c_m\,x^m\right)
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& \mu_n\,\mu_{n'}\, a^2\,\exp\left(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2}\right)
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths
$\lambda_m$ and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.  The exponentiation of
the polynomial in ensures that $f(\alpha)$ is everywhere positive.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise. The
multiplication by $\mu_n\, \mu_{n'}$ in the kernel function signifies
signifies that it is defined in \emph{fractional} terms.

\section{Data and Experimental Setup}

We are now left to demonstrate that, by including photometry as well
as spectroscopy, it is possible to infer the parameters of interest
$\theta$ while marginalizing over the calibration parameters $\alpha$.
We will apply the methodology for spectrophotometric calibration
described above to both \emph{mock} and \emph{real} optical spectra of
Galactic globular clusters.  Spectra of globular clusters provide a
good test case, as they are less complex than galaxy spectra (for
which it is necessary to model the star formation history) but are by
no means trivial. In this way the key features of the calibration
model and methodology are not obscured.  Additionally, information
about the physical parameters of the cluster is available from
resolved stellar CMDs, providing an important check on our inference
methodology and especially our models for star cluster spectra.  However,
our methodology is applicable any spectral model: stars, galaxies or
ISM emission or any object which can be modeled.

@@

Optical photometry is available from \citet{vanderbeke14} based on
SDSS for objects in the northern hemisphere and on data from CTIO for
southern hemisphere objects.  Briefly, total fluxes within the half-light radii
of the clusters \citep{harris96} were calculated, with small
corrections for stars unlikely to be cluster members based on their
positions in the CMD. For this data the formal Poisson uncertainties
are typically exceedingly small.  The dominant source of uncertainty
comes from the photometric calibration and from uncertainties in
cluster membership.

Near-Infrared photometry is available from \citet{cohen07}, who fit
King profiles \citet{king62} to 2MASS data in the $J$, $H$, and $K_s$
bands.  To produce NIR photometry consistent with the optical we
integrated the King profiles to the half-light radii used by
\citet{vanderbeke14}.

@@

The spectrum of one solar mass of stars that formed at lookback time
$t$ with metallicity $Z$ is given by $s(\lambda, t, Z [, \beta])$.
Here the additional parameters $\beta$ may include details of the
population synthesis model, such as the IMF slope and adjustments to
the stellar spectra or stellar evolutionary tracks. We will assume
$\beta$ to be fixed for the purposes of this exercise, with no loss of
generality but significant savings in computational time.  We will
also fix the shape of the attenuation curve, $k_\lambda $, to the
Milky way average of \citet{CCM89} with $R_V=3.1$ and assume that
$d_L$ is known to high precision from independent methods and can
therefore be treated as fixed.  In practice, $d_L$ is perfectly
degenerate with $m_*$. We calculate $s(\lambda, t, Z)$ using the
Flexible Stellar Population Synthesis code \citep[FSPS][]{fsps}.  We
use the Padova 2007 isochrones \citep{girardi00, bertelli94, marigo07}
and the MILES stellar spectral library \citep{miles_I, miles_II,
miles_III}.

@@

\bibitem[Cohen et al.(2007)]{cohen07} Cohen, J.~G., Hsieh, S., 
Metchev, S., Djorgovski, S.~G., \& Malkan, M.\ 2007, \aj, 133, 99

@@

\bibitem[Harris(1996)]{harris96} Harris, W.~E.\ 1996, \aj, 112, 
1487

\bibitem[King(1962)]{king62} King, I.\ 1962, \aj, 67, 471

\bibitem[Koleva et al.(2009)]{koleva09} Koleva, M., 
Prugniel, P., Bouchard, A., \& Wu, Y.\ 2009, \aap, 501, 1269

\bibitem[Lee et al.(2011)]{lee11} Lee, H., Kashyap, V.~L., 
van Dyk, D.~A., et al.\ 2011, \apj, 731, 126

@@

\bibitem[Vanderbeke et al.(2014)]{vanderbeke14} Vanderbeke, J., 
West, M.~J., De Propris, R., et al.\ 2014, \mnras, 437, 1725
